---
title: "Chi-Square (and Measures of Association)"
author: "Colin Hannigan"
format: html
editor: visual
---

### Agenda:

1.  [Install Packages and Libraries](#sec-packages-libraries)
2.  [Remember: `crosstabC()` with the `chisq=T` Argument](#sec-remember-crosstabc-with-the-chisq-argument)
3.  [Application: Making Dummies Using the `ifelse()` Function](#sec-application-making-dummies-using-the-ifelse-function)
4.  [Adding Measures of Association Tests](#sec-adding-measures-of-association-tests)
    a.  [Using the `oii` Package](#sec-using-the-oii-package)
5.  [Advanced Material](#sec-advanced-material)

### 1. Install Packages and Libraries: {#sec-packages-libraries}

#### New packages to install:

If you do not already have these packages, paste them manually into the console and install them.

```{r}
# install.packages("oii")
```

-   [oii](https://cran.r-project.org/web/packages/oii/index.html){target="_blank"} provides additional tests for categorical data in crosstabs.

#### Load libraries!

```{r}
#| message: false
#| warning: false
library(RCPA3)
library(oii)
```

### 2. Remember: `crosstabC()` with the `chisq=T` Argument {#sec-remember-crosstabc-with-the-chisq-argument}

```{r}
crosstabC(dv=nes$biden.honest, iv=nes$religion.group, compact=T, plot.response = "all", chisq = T)
```

There is a significant relationship between R's religious group and the extent to which R believes that the trait "honest" describes President Biden.

```{r}
crosstabC(dv=nes$trump.honest, iv=nes$religion.group, compact=T, plot.response = "all", chisq = T)
```

There is also a significant relationship between R's religious group and the extent to which R believes that the trait "honest" describes President Trump.

BUT the crosstab outputs and the bar charts are, like, a lot to process...

What if we just wanted to look at one or two specific groups as dummy variables?

### 3. Application: Making Dummies Using the `ifelse()` Function {#sec-application-making-dummies-using-the-ifelse-function}

Let's make a variable called "evangelical" to be 1 if `nes$religion.group` is "2. Evangelical Protestant":

```{r}
evangelical<- ifelse(test = nes$religion.group=="2. Evangelical Protestant", "Evangelical", "Not Evangelical")
```

```{r}
evangelical<- as.factor(evangelical) # convert evangelical to a factor to make sure R treats it right
```

Let's make another variable called "catholic" to be 1 if nes\$religion.group is "5. Roman Catholic"

```{r}
catholic<- ifelse(test = nes$religion.group=="5. Roman Catholic", "Catholic", "Not Catholic")
```

```{r}
catholic<- as.factor(catholic) # quick conversion to a factor variable...
```

Now, we can run some crosstabs with chi-square! Let's skip the plots.

```{r}
crosstabC(dv=nes$biden.honest, iv=evangelical, compact=T, chisq = T, plot = F)
```

Do Evangelicals think Biden is honest? Evangelicals are less likely to think Biden is honest (and it's significant).

```{r}
crosstabC(dv=nes$trump.honest, iv=evangelical, compact=T, chisq = T, plot = F)
```

What about Trump? Evangelicals are more likely to think Trump is honest (and it's significant).

```{r}
crosstabC(dv=nes$biden.honest, iv=catholic, compact=T, chisq = T, plot = F)
```

Do Catholics have the same patterns? Not likely. The relationship between being Catholic or not and the extent to which the word "honest" describes Biden is not significant.

```{r}
crosstabC(dv=nes$trump.honest, iv=catholic, compact=T, chisq = T, plot = F)
```

But maybe Catholicism influences belief that Trump is honest? Nah. Still not a significant relationship.

### 4. Adding Measures of Association Tests {#sec-adding-measures-of-association-tests}

```{r}
crosstabC(dv=nes$biden.honest, iv=evangelical, compact=T, chisq = T, plot = F, lambda = T, somers = T, cramers = T)
```

We can add measures of association for Somers' *d*, Cramer's *V*, and Lambda ($\lambda$) directly to our `crosstabC()` chi-square test.

But we're missing some important measures (Gamma and Tau-b)!

#### Using the `oii` Package {#sec-using-the-oii-package}

```{r}
association.measures(y=nes$biden.honest, x=evangelical)
```

Now we can test the other measures of association between the variables in our `crosstabC()` chi-square model.

But we have to be careful to specify x (the IV) and y (the DV) because some measures are asymmetrical!

Note: the `oii` measures of association don't include lambda. I'm okay with it because lambda is mid at best, anyway.

### 5. Advanced Material: {#sec-advanced-material}
