---
title: "Correlation Analysis"
author: "Colin Hannigan"
format: html
editor: visual
---

### Agenda:

1.  [Install Packages and Libraries](#sec-packages-libraries)
2.  [Pearson Correlations](#sec-pearson-correlations)
3.  [Adding p-values to Correlation Output](#sec-adding-p-values-to-correlation-output)
4.  [Adding Scatterplots to Correlation Output](#sec-adding-scatterplots-to-correlation-output)
5.  [Correlations Among Categorical Variables](#sec-correlations-among-categorical-variables)
6.  [Multivariate Correlations](#sec-multivariate-correlations)
7.  [Advanced Material](#sec-advanced-material)

### 1. Install Packages and Libraries: {#sec-packages-libraries}

#### New packages to install:

No new packages to install this time!

#### Load libraries!

```{r}
#| message: false
#| warning: false
library(RCPA3)
```

### 2. Pearson Correlations {#sec-pearson-correlations}

We'll use the `correlateC()` function from the `RCPA3` package to test how strongly continuous/interval variables are associated with one another. The `correlateC()` function requires us to pass a list of variables using the `c()` subfunction. For example, if we wanted to test the correlation between a respondents' feelings toward unions and their feelings toward Congress, it would look like this:

```{r}
correlateC(c(nes$ft.unions, nes$ft.congress))
```

The function produces a correlation of 0.26. How strong is that, based on our rules of the road for evaluating correlation coefficients?

### 3. Adding p-values to Correlation Output {#sec-adding-p-values-to-correlation-output}

Because a Pearson's *r* correlation is technically a hypothesis test, we could also generate evaluate whether the association is statistically significant by simply including the `stats = T` optional argument in the code.

But here's a big **disclaimer**: social scientists rarely do that. We're generally interested in correlations as a sort of descriptive statistic, or a measure of strength in a relationship, not a standalone hypothesis test. Correlation is a helpful starting point or backup test, but it's pretty weak on its own as a test of significance.

```{r}
correlateC(c(nes$ft.unions,nes$ft.congress), stats = T)
```

### 4. Adding Scatterplots to Correlation Output {#sec-adding-scatterplots-to-correlation-output}

However, it definitely helps us understand the distribution of our data and how *strongly* variables are associated. We can also produce a scatterplot of the data to visualize the clusters/patterns in the data. All we need is the `plot = T` optional argument:

```{r}
correlateC(c(nes$ft.unions,nes$ft.congress), plot = T)
```

It's important to note that, in large datasets with strong correlations, we're likely to have overlapping points in the scatterplot that we can't see clearly (like this one). But the `correlateC()` function accounts for that by "jittering" the points, shifting observations a tiny amount to make sure that we can see everything, despite any overlap. The `jitter` option is set to `TRUE` by default. But we can also turn it off if we have to...

```{r}
correlateC(c(nes$ft.unions,nes$ft.congress), plot = T, jitter=F)
```

### 5. Correlations Among Categorical Variables {#sec-correlations-among-categorical-variables}

The `correlateC` function is best suited for continuous/interval IVs *and* DVs. In fact, Pearson's *r* correlation assumes numeric variables explicitly. 

Let's try to include a categorical variable (i.e., a factor variable in R), like `nes$partyid3`, for example:

```{r}
#| error: true
correlateC(c(nes$ft.unions, nes$partyid3), plot = T)
```

Right away, we get an error when we try to correlate factor variable with a continuous variable. That's because of the way R doesn't know to line up categories (i.e., factor variables) against numbers (i.e., numeric variables) in this function and because it would violate the numeric assumption behind Pearson's *r* correlation.

**With great caution, though, we can sometimes use a workaround...**

If we create a copy of our categorical variable that treats it `as.numeric()` instead of as a factor, we can make the correlation run!

```{r}
partyid3_num <- as.numeric(nes$partyid3)
correlateC(c(nes$ft.unions, partyid3_num), plot = T)
```

But this can be problematic for a couple of reasons. For one thing, while we get a clear, interpretable coefficient, the scatterplot isn't as helpful or intuitive. But more importantly, we're making a massive (and sometimes incorrect!) assumption that the numbers assigned to our categories have any meaning apart from simple differentiation. While it makes sense for a dichotomous variable (which can only take on values of 1 or 0), and it could be assumed that *some* ordinal variables will behave linearly, nominal variables are a crapshoot because the numeric values could be in any order.

So, when we use this method to force a correlation involving categorical variables to get a general sense of a relationship, we should be cautious about making any strong inferences.

### 6. Multivariate Correlations {#sec-multivariate-correlations}

We can also correlate multiple variables at once, but `correlateC()` is designed to produce *bivariate* correlations, so we'll get a table of coefficients for each *pair* of variables in our list...

```{r}
correlateC(c(nes$ft.unions, nes$ft.congress, nes$ft.obama, partyid3_num), plot = T, jitter=F)
```
... and a metric ton of plots that are difficult to read. 

So, I recommend using `plot = F` to turn off the plots and just focus on the clear, interpretable coefficients. 

```{r}
correlateC(c(nes$ft.unions, nes$ft.congress, nes$ft.obama, partyid3_num), plot = F, jitter=F)
```

### 7. Advanced Material {#sec-advanced-material}
