---
title: "Logistic Regression (Logit)"
author: "Colin Hannigan"
format: html
editor: visual
---

### Agenda:

1.  [Install Packages and Libraries](#sec-packages-libraries)
2.  [Logistic Regression with `logregC()`](#sec-logistic-regression-with-logregc)
    a.  [Getting Model Statistics](#sec-getting-model-statistics)
    b.  [Classification Tables](#sec-classification-tables)
3.  [Dealing with Ordinal IVs in Logit Models](#sec-dealing-with-ordinal-ivs-in-logit-models)
4.  [Interpreting Partial Effects in Logit](#sec-interpreting-partial-effects-in-logit)
    a.  [Odds Ratios](#sec-odds-ratios)
5.  [Logistic Regression with `glm()`](#sec-logistic-regression-with-glm)
6.  [Readying `glm()` for Interpretation](#sec-readying-glm-for-interpretation)
7.  [Advanced Material](#sec-advanced-material)
    a.  [Manually Calculating Model Stats for `glm()` Models](#sec-manually-calculating-model-stats-for-glm-models)

### 1. Install Packages and Libraries: {#sec-packages-libraries}

#### New packages to install:

No new packages to install this time!

#### Load libraries!

```{r}
#| message: false
#| warning: false
library(RCPA3)
```

### 2. Logistic Regression with `logregC()` {#sec-logistic-regression-with-logregc}

Requirements:

-   DV: Dichotomous
-   IV: Continuous/Interval (bivariate); any (multivariate)

Just like `regC()` for OLS linear regression, `RCPA3` gives us `logregC()` for logistic regression:

```{r}
logregC(formula = world$vdem.2cat ~ world$gdp.percap.5cat + world$frac.eth + world$pop.urban)
```

See how we get p-values for the partial effects (the coefficients/estimates for each individual IV), but not for the model as a whole?

#### Getting Model Statistics {#sec-getting-model-statistics}

To get model statistics, we have to add an optional argument: `fit.stats = T` gives us the model test statistic, which is a special (but different) kind of Chi-square test that runs in the background, and the p-value.

```{r}
logregC(formula = world$vdem.2cat ~ world$gdp.percap.5cat + world$frac.eth + world$pop.urban, fit.stats=T)
```

There are also some other stats that aim to evaluate the "goodness of fit" of the logit model: McFadden's R2, Cox & Snell Index, and Nagelkerke Index are all "Pseudo" R-Squared measures.

Choosing the appropriate Pseudo R-Square measure can be confusing because there are several, but the norm among most social scientists is to rely on **McFadden's R2**.

What makes it a "*Pseudo*" R-Squared is that it attempts to replicate a regular R-Squared (from OLS linear regression) but in a non-linear model. In other words, all of them are kind of imperfect approximations. Even though McFadden's R2 gives us a better general sense of the explanatory power of our model compared to a model without any IVs, it's not a direct analog to a percentage of the variation in the DV explained by the model.

#### Classification Tables {#sec-classification-tables}

That's where a classification table (or a "confusion matrix") comes in. We can also include the `pre = T` argument in the `logregC()` function to get that table and a measure of the Proportional Reduction in Error (PRE), just like we learned about with some of the Chi-square measures of association.

```{r}
logregC(formula = world$vdem.2cat ~ world$gdp.percap.5cat + world$frac.eth + world$pop.urban, fit.stats=T, pre = T)
```

See how there are 44 cases in the upper left box, representing observations that are both *observed* 0's (Autocracy) AND *predicted* 0's? In the middle box, there 62 cases of *observed* 1's (Democracy) that are also *predicted* 1's. These are the cases that our model predicted correctly.

If we add up those two numbers on the diagonal (44 + 62) and divide by the total number of cases (150, from the bottom right box), we get a proportion of correct predictions of the DV.

```{r}
(44+62)/150
```

When we do some mental math to convert to a percentage, we're explaining like 70.66667% of the variation in the DV with our model. Could be better, but not bad!

### 3. Dealing with Ordinal IVs in Logit Models {#sec-dealing-with-ordinal-ivs-in-logit-models}

Notice how the `gdp.percap.5cat` variable is repeated and has weird suffixes? R is trying to fit the ordered factor variable to different polynomial functions to see which one makes the most sense.

-   **.L =** Linear model (treats the effect of the IV as usual, like slope \* X...)
-   **.Q =** Quadratic model (models the IV as slope \* X\^2)
-   **.C =** Cubic model (models the IV as slope \* X\^3)
-   **\^4 =** Quartic model (models the IV as slope \* x\^4)
-   ...and so on for as many categories as there are in the variable, minus 1...

This happens because categorical variables (even ordinal ones) don't necessarily have a linear or consistent effect on the DV at different levels. R tries to find the best "shape" of the IV-DV relationship

In our case, because `world$gdp.percap.5cat.L` is the only significant one of the 4 options, it's likely that the effect of GDP per capita on the probability of being a democracy is fairly linear and consistent.

But there's a less math-y way to understand the relationship: When we treat it like it's unordered (i.e., a nominal variable), R just converts it to a set of dummies, which we can interpret as usual.

We can do that using the `factor()` function with the `unordered = T` option directly in the model...

```{r}
logregC(formula = world$vdem.2cat ~ factor(world$gdp.percap.5cat, ordered = F) + world$frac.eth + world$pop.urban, fit.stats=T, pre = T)
```

...or we can create a new variable to include in the model instead, which can lead to a cleaner model to interpret:

```{r}
gdppc.5cat.nom_ <- factor(world$gdp.percap.5cat, ordered = F)

logregC(formula = world$vdem.2cat ~ gdppc.5cat.nom_ + world$frac.eth + world$pop.urban, fit.stats=T, pre = T)
```

But remember that we'll have to consider the excluded base category, too! We can always change that using the `relevel()` function, too.

### 4. Interpreting Partial Effects in Logit {#sec-interpreting-partial-effects-in-logit}

Because logistic regression is not a linear model, we can't use the same simple vocabulary to interpret the coefficients as we can with linear regression. In fact, apart from direction, the logit coefficients are **not directly interpretable**!

#### Odds Ratios {#sec-odds-ratios}

We can get a little closer to direct interpretation by focusing on the odds ratios instead of the log odds that are produced by default.

To do this, we just add the `orci = T` option to the `logregC()` function:

```{r}
logregC(formula = world$vdem.2cat ~ gdppc.5cat.nom_ + world$frac.eth + world$pop.urban, fit.stats=T, pre = T, orci = T)
```

When we interpret odds ratios, anything above 1 is an *increase* in the odds of being in the higher category (Democracy) and anything below 1 is a *decrease* in the odds of being in the higher category (Democracy). Then we interpret the the

For example, in our model:

-   If pop.urban were significant, a one-unit increase in a country's urban population rate would decrease the odds of a country being a democracy by a factor of 0.997 (a 0.3% decrease in odds).
-   When a state has a GDP per capita of `$45k or more`, its odds of being a democracy significantly increase by a factor of 9.212 (a 821.2% increase in odds).

### 5. Logistic Regression with `glm()`{#sec-logistic-regression-with-glm}

Because `logregC()` is specific to `RCPA3`, it's sometimes helpful to have a base R function to perform a logistic regression. That's where `glm()` comes in. But there are a few common pitfalls to sort through...

Let's start with a basic `glm()` model:

```{r}
#| error: true

glm(states$deathpen.status ~ states$south + states$attend.pct + states$public.conservative + states$gun.dealers)
```

**Error Number 1: DV Problems?**

See how we get an error message? `NA/NaN/Inf in 'y'` is telling us about missing and infinite (i.e., impossible to compute) values in the DV. In other words, there's a problem with our DV and how it's measured.

So, let's check the the DV.

```{r}
attributes(states$deathpen.status)
```

It's a factor and there are only two categories, so it's dichotomous, which is what we need. But maybe the model is better equipped to deal with numbers instead of categories?

We could try converting the dichotomous factor to numeric: either 1 or 2.

```{r}
#| error: true

states$deathpen.status_num <- as.numeric(states$deathpen.status)

glm(deathpen.status_num ~ states$south + states$attend.pct + states$public.conservative + states$gun.dealers)
```

**Error Number 2: Variable Specification Problems?**

Not found?? We definitely created the variable, so this error is confusing. But notice that it doesn't show up in our Environment, either. Carefully re-reading the code, it's clear that we added it to the states dataset, but we didn't specify that dataset in our `glm()` function.

Let's specify the states dataset and try again.

```{r}
glm(states$deathpen.status_num ~ states$south + states$attend.pct + states$public.conservative + states$gun.dealers)
```

It runs this time! But even without throwing an error, **the model is still incorrect**...

**Error Number 3: Logical Problems.**

The default `glm()` method assumes a "Gaussian" distribution, which tries to fit a linear model, assuming normality. In other words, the default is OLS. But we have a dichotomous DV, so a *linear* model doesn't make sense.

Instead, we have to tell `glm()` what type of link/distribution to use when running our model. So, we can go back to the same dichotomous factor DV, but we specify that we want a `binomial` link (i.e., the model has *two* outcomes).

```{r}
glm(states$deathpen.status ~ states$south + states$attend.pct + states$public.conservative + states$gun.dealers, family = binomial)
```

Adding the `family = binomial` argument is exactly what we were missing.

### 6. Readying `glm()` for Interpretation {#sec-readying-glm-for-interpretation}

We'll store the working `glm()` model in an object (`logit1`) so we can view the model summary and reference than object later.

```{r}
logit1 <- glm(states$deathpen.status ~ states$south + states$attend.pct + states$public.conservative + states$gun.dealers, family = binomial)

summary(logit1)
```

While this model will come in handy a little bit later when we present the model results, we don't have model fit statistics, which presents a hurdle with a more involved solution.

But for now, one option is to run both a `glm()` to store and report the model and run a `logregC()` to estimate the model stats and report those separately.

### 7. Advanced Material

#### Manually Calculating Model Stats for `glm()` Models {#sec-manually-calculating-model-stats-for-glm-models}

We'll start by running our fully specified model (`logit1` just like before), but then we'll run another model: the *null model* (`logit0` in this case), which is the same DV but without any of the IVs to help predict it. The null model acts as a baseline for comparison to our model.

```{r}
logit1 <- glm(states$deathpen.status ~ states$south + states$attend.pct + states$public.conservative + states$gun.dealers, family = binomial)

logit0 <- glm(states$deathpen.status ~ 1, family = binomial)
```

Then, to get the p-value of the model as whole, we have to compare the log likelihood of our model to the log likelihood of the null model using a **likelihood ratio test**:

```{r}
pvalue <- anova(logit0, logit1, test = "Chisq")
```

Wait. ANOVA? Chi-square? Again??

Yep.

Earlier on, when we looked at bivariate tests, ANOVA helped us compare means and Chi-square helped us analyze crosstabs with categorical variables. With logistic regression, we're asking different questions:

-   Does a model *with* IVs fit the data significantly better than a model *without* IVs?

So, R is just using the `anova()` function to compare the log likelihoods of both those models to produce a likelihood ratio test statistic, then using a Chi-square test to determine whether the improvement in model fit is large enough to be significant.

Next, we can calculate McFadden's pseudo R-squared to see how much the full model improves the log-likelihood compare to the null model. So, we specify the `logLik()` parameters stored in each model:

```{r}
pseudoR2 <- as.numeric(1 - (logLik(logit1) / logLik(logit0)))
```

Similarly, we can extract the coefficient parameters from the stored model, so we can exponentiate them into Odds Ratios:

```{r}
oddsRatios <- exp(coef(logit1))
```

Finally, we just call each of the objects to print them to the console and see those values. The `pvalue` line will print the results of the ANOVA we ran above, which includes "Pr(\>Chi)" that we can report. Then, `pseudoR2` will report a number for our McFadden's pseudo R-squared, and the `oddsRatios` will produce a table of each of the IVs and their Odds Ratios which we can report after comparing to the original table and seeing whether the partial effects there are significant.

```{r}
pvalue

pseudoR2

oddsRatios
```

Now, we can run a `logregC()` and see that the values we just calculated are the same (or close enough, given rounding decisions)!

```{r}
logregC(states$deathpen.status ~ states$south + states$attend.pct + states$public.conservative + states$gun.dealers, fit.stats = T, pre=T, orci = T)
```
