---
title: "t-tests and ANOVAs"
author: "Colin Hannigan"
format: html
editor: visual
---

### Agenda:

1.  [Install Packages and Libraries](#sec-packages-libraries)
2.  [Remember `compmeansC()`](#sec-2.-remember-compmeansc)
3.  [Testing Differences in Proportions](#sec-3.-testing-differences-in-proportions)
4.  [Testing Differences in Means](#sec-4.-testing-differences-in-means)
    a.  [One-Sample t-test](#sec-one-sample-t-test)
    b.  [Two-Sample t-test](#sec-two-sample-t-test)
    c.  [Paired Samples t-test](#sec-paired-samples-t-test)
5.  [Analysis of Variance Tests (ANOVAs)](#sec-5.-analysis-of-variance-tests-anovas)
    a.  [Tukey Test: Post-Hoc tests with ANOVAs](#sec-tukey-test-post-hoc-tests-with-anovas)
6.  [Advanced Material](#sec-6.-advanced-material)

### 1. Install Packages and Libraries: {#sec-packages-libraries}

#### New packages to install:

No new packages to install this time!

#### Load libraries!

```{r}
#| message: false
#| warning: false
library(RCPA3)
```

### 2. Remember `compmeansC()` {#sec-2.-remember-compmeansc}

```{r}
compmeansC(world$gdp.percap, world$vdem.4cat)
```

We can see the mean GDP per capita for each of the 4 regime type categories. But is the difference **significant**?? The `compmeansC()` function can't tell us the significance; it's more descriptive.

### 3. Testing Differences in Proportions {#sec-3.-testing-differences-in-proportions}

Instead, we can do a "one-sample test" of proportions to see if the proportion of cases for a given category is significantly different from our hypothesized proportion:

```{r}
testpropsC(world$vdem.4cat, .25, response = "1. Liberal democracy")
```

Is the proportion of Liberal Democracies significantly different from the actual proportion of Liberal Democracies in the world?

### [4. Testing Differences in Means](slides/ttestsAnova.pdf#page=7) {#sec-4.-testing-differences-in-means}

We use what's called a "t-test" to test when means are significantly different

-   DV: Continuous/Interval
-   IV: Dichotomous

**There are 3 types of t-tests:**

1.  One-sample t-test
2.  Two-sample t-test
3.  Paired-sample t-test

#### One-Sample t-test {#sec-one-sample-t-test}

One-sample t-test: Is the mean of Y significantly different from a specific value?

```{r}
testmeansC(states$vep20.turnout, 50)
```

Was mean voter turnout by state in 2020 significantly different from 50%?

#### Two-Sample t-test {#sec-two-sample-t-test}

Two-sample t-test: Is the mean of Y (when the IV=0) significantly different from the mean of Y (when the IV=1)?

```{r}
testmeansC(dv= states$vep20.turnout, iv=states$term.limits)
```

Was mean voter turnout in 2020 among states with term limits significantly different from turnout among states without term limits?

#### Paired samples t-test {#sec-paired-samples-t-test}

Paired-samples t-test: Is the mean of Y1 significantly different from the mean of a related variable, Y2? (E.g., Pre-test vs. Post-test). Even though we don't specify an X variable, the dichotomous IV is ***implicit*** here.

Instead of each of the Y variables being unrelated columns, like this:

| ID            | Y1  | Y2  |
|---------------|-----|-----|
| Observation 1 | 111 | 222 |
| Observation 2 | 333 | 444 |
| Observation 3 | 555 | 666 |

Imagine the data is rearranged to look like this:

| ID            | Value | Which Y? |
|---------------|-------|----------|
| Observation 1 | 111   | Y1       |
| Observation 1 | 222   | Y2       |
| Observation 2 | 333   | Y1       |
| Observation 2 | 444   | Y2       |
| Observation 3 | 555   | Y1       |
| Observation 3 | 666   | Y2       |

```{r}
testmeansC(states$vep20.turnout, states$vep16.turnout, paired = T)
```

Was mean voter turnout in 2020 significantly different from mean voter turnout in 2016?

```{r}
#| message: false

load("datasets/testScoresData.RData")
```

```{r}
testmeansC(testScores$score.pre, testScores$score.post, paired = T)
```

Was mean pre-test score significantly different from mean post-test score for that fake class data I showed you?

But what if we have **multiple categories in the IV**??

### [5. Analysis of Variance Tests (ANOVAs)](slides/ttestsAnova.pdf#page=17) {#sec-5.-analysis-of-variance-tests-anovas}

T-tests require an explicit or implicit *dichotomous* IV. So, when we have multiple categories we need a different test:

ANOVAs (Analysis of Variance):

-   DV: Continuous/Interval
-   IV: Nominal or Ordinal

```{r}
compmeansC(world$gdp.percap, world$vdem.4cat, anova = T)
```

We can add an ANOVA test option to the `compmeansC()` function but it's not great. The `anova = T` option tells us whether the relationship *as a whole* is significant, but we don't know what's driving that significance because it doesn't break things down by the individual categories of the IV.

That's where `aov()` makes a difference...

```{r}
anova.model1 <- aov(world$gdp.percap~world$vdem.4cat)
```

The `aov()` function runs an ANOVA model, too. And, importantly, we can save this model as an object called `anova.model1` (or whatever) so we can call it again. Notice that we don't include a comma, though: the syntax is DV\~IV (with the tilde).

Let's see the model:

```{r}
summary(anova.model1)
```

Getting a summary of the anova model object gives us the p-value of the model as a whole (just like `compmeansC()` with the `anova = T` option). But since we've stored it as an object, we have a additional opportunities...

#### Tukey Test: Post-Hoc tests with ANOVAs {#sec-tukey-test-post-hoc-tests-with-anovas}

Here's the cool part of creating a model object with `aov()`: we can run a "post-hoc test" (a test after the fact) on it to see how *each* of the IV categories relates to *each* of the DV categories.

This is the Tukey HSD ("Honestly Significant Difference") test.

```{r}
TukeyHSD(anova.model1)
```

The `TukeyHSD()` test output is key to our model interpretation. It lists each category comparison within the IV (e.g., "2. Electoral democracy-1. Liberal democracy") and provides some stats: the difference in mean DV `world$gdp.percap` between them, the lower (`lwr`) and upper (`upr`) bounds of the confidence interval, and an adjusted p-value (`p adj`) for each individual comparison.

### 6. Advanced Material {#sec-6.-advanced-material}
